---
title: "Lab 1: NYC Flights"
author: "Aaron C Cochran"
date: "October 5, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nycflights13)
library(tidyverse)
data(flights)
```

# Objectives

In this first lab you will:

- use an established r dataset from `nycflights13`
- explore the dataset 
- create graphical summaries

Most of this lab comes directly from the R for Data Science book by Hadley Wickham and Garrett Grolemund, available at <http://r4ds.had.co.nz/>.

## Load Data

Load the data and check out the structure of the dataset.

```{r demo-load-data, eval=TRUE, echo=TRUE}
library(nycflights13)
data(flights)
head(flights)
```


### Questions

How many rows are in the dataset?

How many variables?

What type (class) of variable is the time_hour variable?

## Exploring the data

Let's dig into the dataset and explore it.


```{r dplyr-unique, eval=TRUE, echo=TRUE}

# using dplyr's n_distinct is the same as length(unique(x)) from base R
# but faster and easier to use
n_distinct(flights$tailnum)

# how many origin airports in NY?
n_distinct(flights$origin)

# what are the origin airports?
levels(as.factor(flights$origin))

# how many destinations? 
# (note: way more, probably don't want to list them all)
n_distinct(flights$dest)

```

How many total flights departed from just JFK?  This will use the `filter` command from `dplyr`. 

> Note the use of the double equals sign == which is the equivalent of "is equal to"

```{r dplyr-filter}

flights %>% 
  filter(origin == 'JFK') %>% 
  nrow() # nrow() gives the number of rows

```

What was the average departure delay? What is the standard deviation?

```{r dplyr-summary}

summarize(flights, mean = mean(dep_delay, na.rm=T), sd = sd(dep_delay, na.rm=T)) 
# note the na.rm=T argument. This tells R to ignore (remove) NA
# values when calculating the mean and standard deviation

```


### Questions

1. How many flights departed from LaGuardia (LGA) for Portland (PDX)

> Hint: You can combine filtered terms with & (and) or | (or). 

```{r q2-1, include=TRUE}

flights %>% 
  filter(origin == 'JFK' & dest == 'PDX') %>% 
  nrow()


```

2. What was the average `air_time` for these flights?

```{r q2-2, include=TRUE}

flights %>% 
  filter(origin == 'JFK' & dest == 'PDX') %>% 
  summarise(mean_air_time = mean(air_time, na.rm=T), sd_air_time = sd(air_time, na.rm=T))
  
```

3. How many flights from each airport happened in July?

```{r q2-3, include=TRUE}

flights %>% 
  filter(month == 7) %>% 
  group_by(month, origin) %>% 
  summarise(count = n())

```


## Visualizing the data

#### geom_histogram()

Using `ggplot2` let's create some graphical summaries of the data. 

```{r plot-1}

flights %>% filter(origin == 'JFK' & dest == 'PDX') %>%  
  # note that ggplot2 doesn't use the %>% pipe operator
  # it was written before that was adopted
  # so it still uses a + sign. 
  ggplot() + 
  # define the geom
  geom_histogram(aes(x=factor(month)), stat='count') +
  # x-axis title
  scale_x_discrete("Month") +
  # y-axis title
  scale_y_continuous('Flights') +
  # graph title
  ggtitle('Flights from JFK to PDX, by month') 

```

#### facet_wrap()


```{r plot-2}
# all airports to PDX as a facet
flights %>% filter(dest=='PDX') %>% 
  # note that LGA does not fly to PDX so it is automatically filtered out
  ggplot() +
  geom_histogram(aes(x=factor(month)), stat='count') +
  # facet_wrap splits the graphs up by the specified variable
  facet_wrap(~origin) +
  # title and labels as before
  scale_x_discrete('Month') +
  ggtitle('Flights from NY Airports to PDX, by month')
  
```

#### geom_density()

`geom_density()` creates density plots, which can be easily overlayed to show distributions between groups. 

Example:

Do certain carriers fly more long or short routes?  Let's look just at flights leaving `JFK`, by carrier, with respect to total `air_time`. For this, we'll need to use another dataset in the `nycflights13` package that contains the names of the airlines, instead of just the carrier codes. This will require a `left_join` (SQL users will recognize this term) on a common column between the two datasets.


```{r plot-3}
data(airlines) # loaded from nycflights13 package

flights  %>% # start with flight data
  left_join(airlines, by="carrier") %>% # join with airlines data to get names of airlines 
  subset(origin == 'JFK') %>% # only flights leaving from JFK
  ggplot(aes(x=air_time)) +  # generate ggplot object
  geom_density(stat='density', # add density plot geom
               aes(fill=factor(name)), # name is the new column we joined to our data
               alpha = .5) + # alpha sets transparency
  ggtitle('Flight time for JFK departures, by airline') # title
```


### Exercises


1. Create a graph showing flights to `PDX` by month from both `EWR` and `JFK`, but as a dodged bar chart instead of a faceted one. 

> Hint: position = "dodge" is the option for geom_bar()

```{r, include=TRUE}
flights %>% filter(dest=='PDX') %>%
  ggplot() +
  geom_bar(position = "dodge", stat = "count", aes(x=factor(month), fill=origin)) +
  ggtitle('Flights to PDX by month, by origin')
```

2. What types of planes, and how many of each does Jet Blue (`carrier == B6`) fly? 

> Hint: You'll need to join with the `planes` dataset in the `nycflights13` package for this one.

```{r, include=TRUE}
flights %>%
  left_join(planes, by="tailnum") %>% # join with planes by tailnum
  filter(carrier == 'B6') %>% # filtered to just include Jet Blue (B6)
  ggplot() + geom_bar(stat = 'count', aes(x=factor(model))) +
  geom_label(stat='count', aes(label = ..count.., x=factor(model)), vjust=.35) +
  theme(axis.text.x = element_text(angle = 0)) + # Change angle to rotate text
  xlab('Aircraft Model') +
  ggtitle('Jet Blue airplane models, 2013 NYC Flights')
```



